---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# treerabid

*This is a work in progress that will break | change | may get rehomed!*

`treerabid` reconstructs transmission trees using line list data--specifically in the 
context of contact tracing data for canine rabies in Tanzania for the Hampson Lab. 

Based on:
- [Hampson et al. 2009. Transmission Dynamics and Prospects for the Elimination of Canine Rabies.](https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.1000053)
- [Cori et al. 2019. A graph-based evidence synthesis approach to detecting outbreak clusters: An application to dog rabies.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006554)
- Mancy et al. in prep.

## Installation

Install from github with:

``` r
# install.packages("devtools")
devtools::install_github("mrajeev08/treerabid")
```

Dependencies: `data.table`, `foreach`, `doRNG`, `parallel`
Suggests: `ggraph`, `ggplot2`, `igraph`

## Example using `treerabid` + `simrabid`

```{r, message=FALSE}
# Dependencies for simrabid
library(raster)
library(data.table)
library(sf)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(fasterize)
library(lubridate)

# Additional dependencies for treerabid
library(igraph)
library(ggraph)
library(foreach)
library(doRNG)
library(doParallel)

# simrabid & treerabid
library(simrabid) # devtools::install_github("mrajeev08/simrabid")
devtools::load_all()

```


First simulate from rabies IBM using `simrabid`:

```{r sim}
# set up 
sd_shapefile <- st_read(system.file("extdata/sd_shapefile.shp", 
                                    package = "simrabid"))

# 1. set up the space at 1000 m resolution
sd_shapefile$id_col <- 1:nrow(sd_shapefile)
out <- setup_space(shapefile = sd_shapefile, resolution = 1000, id_col = "id_col", 
                   use_fasterize = TRUE)
pop_out <- out
values(pop_out) <- rpois(ncell(pop_out), 20) # fake some population data
pop_out[is.na(out)] <- NA
plot(pop_out)

# 2. set-up simulation framework 
start_up <- setup_sim(start_date = "2002-01-01",
                      apprx_end_date = "2012-01-01", # apprx 10 years
                      days_in_step = 7, # weekly timestep
                      rast = out, 
                      death_rate_annual = 0.48, 
                      birth_rate_annual = 0.52,
                      waning_rate_annual = 1/3,
                      params = list(start_pop = pop_out[]), 
                      by_admin = FALSE)

# 3. Simulate vaccination
vacc_dt <- simrabid::sim_campaigns(locs = 1:75, campaign_prob = 0.7, 
                                   coverage = 0.4, sim_years = 10, 
                                   burn_in_years = 0,
                                   steps_in_year = 52)

# 4. Run the simulation
# see ?simrabid for more details on function arguments
system.time({
  set.seed(1244)
  exe <- simrabid(start_up, start_vacc = 0, I_seeds = 0,
                 vacc_dt = vacc_dt,
                 params = c(list(R0 = 1.1, k = 1, iota = 0.25),
                            param_defaults),
                 days_in_step = 7,
                 observe_fun = beta_detect_monthly,
                 serial_fun = serial_lognorm,
                 dispersal_fun = dispersal_lognorm,
                 secondary_fun = nbinom_constrained,
                 incursion_fun = sim_incursions_pois,
                 movement_fun = sim_movement_continuous,
                 sequential = FALSE, allow_invalid = TRUE,
                 leave_bounds = TRUE, max_tries = 100,
                 summary_fun = use_mget, 
                 track = FALSE,
                 weights = NULL,
                 row_probs = NULL,
                 coverage = TRUE,
                 break_threshold = 0.8, 
                 by_admin = FALSE) 
}
)

# I_dt is the line list
case_dt <- exe$I_dt
head(case_dt)

```

Reconstruct bootstrapped trees (per Hampson et al. 2009) & prune any unlikely case pairs based on the distribution of distances between cases and a pecentile cutoff (see Cori et al):

```{r trees}
# turn time step to dates
case_dt$date <- as_date(duration(case_dt$t_infected, "weeks") + ymd(start_up$start_date))
# construct one tree
ttrees <- 
  boot_trees(id_case = case_dt$id,
             id_biter = 0, # we don't know the progenitors 
             x_coord = case_dt$x_coord,
             y_coord = case_dt$y_coord,
             owned = 0, 
             date_symptoms = case_dt$date,
             days_uncertain = 0,
             use_known_source = FALSE,
             prune = TRUE,
             si_fun = si_gamma1,
             dist_fun = dist_weibull1, 
             params = params_treerabid, 
             cutoff = 0.95,
             N = 1, 
             seed = 105)
ttrees2 <- 
  boot_trees(id_case = case_dt$id,
             id_biter = 0, # we don't know the progenitors 
             x_coord = case_dt$x_coord,
             y_coord = case_dt$y_coord,
             owned = 0, 
             date_symptoms = case_dt$date,
             days_uncertain = 0,
             use_known_source = FALSE,
             prune = TRUE,
             si_fun = si_gamma1,
             dist_fun = dist_weibull1, 
             params = params_treerabid, 
             cutoff = 0.95,
             N = 1, 
             seed = 105)

# Are these reproducible?
identical(ttrees, ttrees2)

# Lets do 100 trees and vizualize them
system.time({
  ttrees <- 
        boot_trees(id_case = case_dt$id,
                   id_biter = 0, # we don't know the progenitors 
                   x_coord = case_dt$x_coord,
                   y_coord = case_dt$y_coord,
                   owned = 0, 
                   date_symptoms = case_dt$date,
                   days_uncertain = 0,
                   exclude_progen = FALSE, 
                   use_known_source = FALSE,
                   prune = TRUE,
                   si_fun = si_gamma1,
                   dist_fun = dist_weibull1, 
                   params = params_treerabid, 
                   cutoff = 0.95,
                   N = 100, 
                   seed = 105)
})

```

## Visualizing trees

We can then visualize the potential links:
```{r}
links_all <- build_all_links(ttrees, N = 100)
links_gr <- graph_from_data_frame(d = data.frame(from = links_all$id_progen, 
                                                 to = links_all$id_case))
E(links_gr)$prob <- links_all$prob
V(links_gr)$membership <- components(links_gr)$membership

# Get rid of the NA links (i.e. differentiating incursions)
links_gr <- delete_vertices(links_gr, names(V(links_gr)) %in% "NA")

set.seed(179)
ggraph(links_gr, layout = "kk") + 
  geom_edge_link0(aes(col = prob), alpha = 0.5) +
  geom_node_point(aes(col = factor(membership)), size = 0.3) +
  scale_color_discrete(guide = "none") +
  scale_edge_color_distiller(direction = 1) +
  theme_graph()
```

Visualize the consensus links and how certain they are:
```{r}

# get the time!
links_consensus <- build_consensus_links(links_all, 
                                         case_dates = case_dt[, .(id_case = id, 
                                                                  symptoms_started = date)])
cons_gr <- get_graph(from = links_consensus$id_progen, 
                     to = links_consensus$id_case, 
                     attrs = case_dt[, .(id_case = id, 
                                         t = 0)])
E(cons_gr)$prob <- links_consensus[!is.na(id_progen)]$prob
V(cons_gr)$membership <- components(cons_gr)$membership

set.seed(145)
# color incursions by membership + alpha = their probability
ggraph(cons_gr, layout="kk") + 
  geom_edge_link0(aes(col = prob), alpha = 0.5) +
  geom_node_point(aes(col = factor(membership)),size = 0.3) +
  scale_edge_color_distiller(direction = 1) +
  scale_color_discrete(guide = "none") +
  theme_graph()

```

Incursions are those that didn't have any potential progenitor within the cutoff 
time & distance. We can see the probability for each case being an incursion (total and 
for those that were assigned as such) (actually this is always one because not any uncertainty in dates!)

```{r, eval=FALSE}
incs_all <- links_all[is.na(id_progen)]

ggplot(incs_all) +
  geom_histogram(aes(x = prob))

```


Compute chains stats on the consensus links:

```{r}
chain_stats <- get_chain_stats(cons_gr)
ggplot(chain_stats) +
  geom_histogram(aes(x = size))
ggplot(chain_stats) +
  geom_histogram(aes(x = length))

```


We can also vizualize the consensus tree (i.e. tree which includes the highest % of consensus links):

```{r}
tree_consensus <- build_consensus_tree(links_consensus, ttrees)
```

# Customize it

Writing your own si and distance distribution functions:
```{r}
# customizing params 
```


Let's pretend that we actually know some of these case pairs from contact tracing data and reconstruct only unknown links:

```{r known}
# using known traced data
```

## Some preliminary work re: estimating detection probabilities

Cori et al. 2019 use analytic expectations of temporal, spatial, genetic distances
between case pairs given a detection probability. 

Working backwards from there, if we can link case pairs to get the 
distribution of distances (genetic | spatial | temporal) between case pairs, can we compare this to the analytical expectations to estimate detection probabilities?

Right now, only one that really works in temporal:

```{r }

# Line list full | 85% | 50% | 20% data & estimate detection 

```

Also the genetic data:

```{r detection}

# Simplified assignment of lineages & snps 

# Current circulating lineages to sample from

# Plus probability of novel lineage being introduced into the Serengeti

# snp distribution between lineages

# mutation rate within lineages (assume that between lineage snp + within lineage snps are additive?) 

# Line list with 25% | 10% | 5% data & estimate detection 

```

And this is using known case links, so really the *perfect* data, i.e. if it doesn't work in this context, it's unlikely to work in real data world. The real test here would be to simulat with known10.714 detection probability, reconstruct transmission tree, and then try to recover the detection probability. 

## Known limitations & future directions

- Speeding up / more efficient parallelization
    - right now it's too mem heavy to use a cluster size larger than 3 cores! 
    - part to parallelize = the data.table joins? or the distance and si functions?
    - make sure to manage data.table threads explicitly through setDTthreads
- Better incorporating uncertainty in contact tracing data to avoid creating loops
  & getting reasonable dates?
- Incorporating uncertainty into location data?
- Using N data streams with N uncertainties & N cutoffs/prob distributions
- Using genetic data (SNPs between cases per Cori | Lineage assignemnts)
- Using partial data (i.e. incorporating genetic data when it is available)
- Use simulations to validate the detection estimates
    - Get the distance | empirical bits working in the detection est
    - Simulate from simrabid + reconstruct bootstrapped trees + 
      estimate detection for each tree & compare to known estimate
    - Combining data streams in this context?
