---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# treerabid

`treerabid` reconstructs transmission trees using line list data--specifically in the 
context of contact tracing data for canine rabies in Tanzania for the Hampson Lab. 

It is still in active development and may get rehomed in the future. Before using, we highly recommend submitting an issue to this repository. 

For the package used in Mancy et al/Lushasi et al.:
[![DOI](https://zenodo.org/badge/361056754.svg)](https://zenodo.org/badge/latestdoi/361056754)

Or install version 1.0 from github:
```
devtools::install_github("mrajeev08/treerabid@v1.0", dependencies = TRUE)
```

Based on:
- [Hampson et al. 2009. Transmission Dynamics and Prospects for the Elimination of Canine Rabies.](https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.1000053)
- [Cori et al. 2019. A graph-based evidence synthesis approach to detecting outbreak clusters: An application to dog rabies.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006554)
- Mancy et al. in prep.

## Installation

Install from github with:

``` r
# install.packages("devtools")
devtools::install_github("mrajeev08/treerabid", dependencies = TRUE)
```

Dependencies: `data.table`, `foreach`, `doRNG`, `parallel`
Suggests: `ggraph`, `ggplot2`, `igraph`

## Example using `treerabid` + `simrabid`

```{r, message=FALSE}
# Dependencies for simrabid
library(raster)
library(data.table)
library(sf)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(fasterize)
library(lubridate)

# Additional dependencies for treerabid
library(igraph)
library(ggraph)
library(foreach)
library(doRNG)
library(doParallel)

# simrabid & treerabid
library(simrabid) # devtools::install_github("mrajeev08/simrabid")
library(treerabid)
```


First simulate from rabies IBM using `simrabid`:

```{r sim}
# set up 
sd_shapefile <- st_read(system.file("extdata/sd_shapefile.shp", 
                                    package = "simrabid"))

# 1. set up the space at 1000 m resolution
sd_shapefile$id_col <- 1:nrow(sd_shapefile)
out <- setup_space(shapefile = sd_shapefile, resolution = 1000, id_col = "id_col", 
                   use_fasterize = TRUE)
pop_out <- out
values(pop_out) <- rpois(ncell(pop_out), 20) # fake some population data
pop_out[is.na(out)] <- NA
plot(pop_out)

# 2. set-up simulation framework 
start_up <- setup_sim(start_date = "2002-01-01",
                      apprx_end_date = "2012-01-01", # apprx 10 years
                      days_in_step = 7, # weekly timestep
                      rast = out, 
                      death_rate_annual = 0.48, 
                      birth_rate_annual = 0.52,
                      waning_rate_annual = 1/3,
                      params = list(start_pop = pop_out[]), 
                      by_admin = FALSE)

# 3. Simulate vaccination
vacc_dt <- simrabid::sim_campaigns(locs = 1:75, campaign_prob = 0.7, 
                                   coverage = 0.4, sim_years = 10, 
                                   burn_in_years = 0,
                                   steps_in_year = 52)

# 4. Run the simulation
# see ?simrabid for more details on function arguments
system.time({
  set.seed(1244)
  exe <- simrabid(start_up, start_vacc = 0, I_seeds = 0,
                 vacc_dt = vacc_dt,
                 params = c(list(R0 = 1.1, k = 1, iota = 0.25),
                            param_defaults),
                 days_in_step = 7,
                 observe_fun = beta_detect_monthly,
                 serial_fun = serial_lognorm,
                 dispersal_fun = dispersal_lognorm,
                 secondary_fun = nbinom_constrained,
                 incursion_fun = sim_incursions_pois,
                 movement_fun = sim_movement_continuous,
                 sequential = FALSE, allow_invalid = TRUE,
                 leave_bounds = TRUE, max_tries = 100,
                 summary_fun = use_mget, 
                 track = FALSE,
                 weights = NULL,
                 row_probs = NULL,
                 coverage = TRUE,
                 break_threshold = 0.8, 
                 by_admin = FALSE) 
}
)

# I_dt is the line list
case_dt <- exe$I_dt
head(case_dt)

```

Reconstruct bootstrapped trees (per Hampson et al. 2009) & prune any unlikely case pairs based on the distribution of distances between cases and a pecentile cutoff (see Cori et al):

```{r trees}
# turn time step to dates
case_dt$date <- as_date(duration(case_dt$t_infected, "weeks") + ymd(start_up$start_date))
# construct one tree
ttrees <- 
  boot_trees(id_case = case_dt$id,
             id_biter = 0, # we don't know the progenitors 
             x_coord = case_dt$x_coord,
             y_coord = case_dt$y_coord,
             owned = 0, 
             date_symptoms = case_dt$date,
             days_uncertain = 0,
             use_known_source = FALSE,
             prune = TRUE,
             si_fun = si_gamma1,
             dist_fun = dist_weibull1, 
             params = params_treerabid, 
             cutoff = 0.95,
             N = 1, 
             seed = 105)
ttrees2 <- 
  boot_trees(id_case = case_dt$id,
             id_biter = 0, # we don't know the progenitors 
             x_coord = case_dt$x_coord,
             y_coord = case_dt$y_coord,
             owned = 0, 
             date_symptoms = case_dt$date,
             days_uncertain = 0,
             use_known_source = FALSE,
             prune = TRUE,
             si_fun = si_gamma1,
             dist_fun = dist_weibull1, 
             params = params_treerabid, 
             cutoff = 0.95,
             N = 1, 
             seed = 105)

# Are these reproducible?
identical(ttrees, ttrees2)

# Lets do 100 trees and vizualize them
system.time({
  ttrees <- 
        boot_trees(id_case = case_dt$id,
                   id_biter = 0, # we don't know the progenitors 
                   x_coord = case_dt$x_coord,
                   y_coord = case_dt$y_coord,
                   owned = 0, 
                   date_symptoms = case_dt$date,
                   days_uncertain = 0,
                   exclude_progen = FALSE, 
                   use_known_source = FALSE,
                   prune = TRUE,
                   si_fun = si_gamma1,
                   dist_fun = dist_weibull1, 
                   params = params_treerabid, 
                   cutoff = 0.95,
                   N = 100, 
                   seed = 105)
})

```

## Visualizing trees

We can then visualize the potential links:
```{r}
links_all <- build_all_links(ttrees, N = 100)
links_gr <- graph_from_data_frame(d = data.frame(from = links_all$id_progen, 
                                                 to = links_all$id_case))
E(links_gr)$prob <- links_all$prob
V(links_gr)$membership <- components(links_gr)$membership

# Get rid of the NA links (i.e. differentiating incursions)
links_gr <- delete_vertices(links_gr, names(V(links_gr)) %in% "NA")

set.seed(179)
ggraph(links_gr, layout = "kk") + 
  geom_edge_link0(aes(col = prob), alpha = 0.5) +
  geom_node_point(aes(col = factor(membership)), size = 0.3) +
  scale_color_discrete(guide = "none") +
  scale_edge_color_distiller(direction = 1) +
  theme_graph()
```

Visualize the consensus links and how certain they are:
```{r}

# get the time!
links_consensus <- build_consensus_links(links_all, 
                                         case_dates = case_dt[, .(id_case = id, 
                                                                  symptoms_started = date)])
cons_gr <- get_graph(from = links_consensus$id_progen, 
                     to = links_consensus$id_case, 
                     attrs = case_dt[, .(id_case = id, 
                                         t = 0)])
E(cons_gr)$prob <- links_consensus[!is.na(id_progen)]$prob
V(cons_gr)$membership <- components(cons_gr)$membership

set.seed(145)
# color incursions by membership + alpha = their probability
ggraph(cons_gr, layout="kk") + 
  geom_edge_link0(aes(col = prob), alpha = 0.5) +
  geom_node_point(aes(col = factor(membership)),size = 0.3) +
  scale_edge_color_distiller(direction = 1) +
  scale_color_discrete(guide = "none") +
  theme_graph()

```

Incursions are those that didn't have any potential progenitor within the cutoff 
time & distance. We can see the probability for each case being an incursion (total and 
for those that were assigned as such) (actually this is always one because not any uncertainty in dates!)

```{r, eval=FALSE}
incs_all <- links_all[is.na(id_progen)]

ggplot(incs_all) +
  geom_histogram(aes(x = prob))

```


Compute chains stats on the consensus links:

```{r}
chain_stats <- get_chain_stats(cons_gr)
ggplot(chain_stats) +
  geom_histogram(aes(x = size))
ggplot(chain_stats) +
  geom_histogram(aes(x = length))

```


We can also vizualize the consensus tree (i.e. tree which includes the highest % of consensus links):

```{r}
tree_consensus <- build_consensus_tree(links_consensus, ttrees, links_all)
```

